---
title: Cross Tabulation
---

Creating contingency tables (cross-tabulations) in R can be done using base R as well as functions from various R packages like `dplyr`, `tidyr`, `janitor`, and `gmodels`. Here's an overview:


### Base R: `table()`

The simplest way to create a contingency table is using `table()` in base R.

```{r}
# Example data
data <- data.frame(
  Gender = c("Male", "Female", "Female", "Male", "Female"),
  Preference = c("A", "B", "A", "B", "A")
)

# Create a contingency table
table(data$Gender, data$Preference)
```


### `dplyr` with `count()`

You can use the `dplyr` package to count combinations of two variables.

```{r}
library(dplyr)

data %>%
  count(Gender, Preference)
```



### tidyr` with `pivot_wider()`

You can create a contingency table using `pivot_wider()` in `tidyr`.

```{r}
library(tidyr)

data %>%
  count(Gender, Preference) %>%
  pivot_wider(names_from = Preference, values_from = n, values_fill = 0)
```



### `janitor` with `tabyl()`

The `janitor` package provides the `tabyl()` function for quick and clean contingency tables.

```{r}
library(janitor)

data %>%
  tabyl(Gender, Preference)
```
