---
title: "Compare: Two Proprotion"
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
here::i_am("stat/prop-compare.qmd")
library(here)
```

```{r}
library(ggplot2)
```


## chi-square goodness of fit test

```{r}
# load iris dataset
dat <- iris

# create size variable
dat$size <- ifelse(dat$Sepal.Length < median(dat$Sepal.Length),
  "small", "big"
)

# show first 5 observations
head(dat, n = 5)
```



```{r}
# chi-square goodness of fit test
test <- chisq.test(table(dat$Species), # observed proportions
  p = c(1 / 3, 1 / 3, 1 / 3) # expected proportions
)

test
```



## Chi-square test vs Fisher's exact test

**Usual rule:** Chi-square test is not appropriate when the expected values in one of the cells of the contingency table is **less than 5**, and in this case the Fisher’s exact test is preferred 


## Fisher's exact test

- [Statsandr Fisher's exact test](https://statsandr.com/blog/fisher-s-exact-test-in-r-independence-test-for-a-small-sample/)

the Fisher’s exact test is used when the sample is small (and in this case the 
p-value is exact and is not an approximation).


```{r}
dat <- data.frame(
  "smoke_no" = c(7, 0),
  "smoke_yes" = c(2, 5),
  row.names = c("Athlete", "Non-athlete"),
  stringsAsFactors = FALSE
)
colnames(dat) <- c("Non-smoker", "Smoker")

dat
```

### Mosaic Plot

```{r}
mosaicplot(dat,
  main = "Mosaic plot",
  color = TRUE
)
```

### Expected frequencies

Fisher’s exact test is used when there is at least one cell in the contingency table of the expected frequencies below 5.

```{r}
chisq.test(dat)$expected
```

### Perform

#### For Contingency Table

```{r}
dat
```


```{r}
test <- fisher.test(dat)
test
```

```{r}
rstatix::fisher_test(dat, detailed = TRUE)
```


#### Long Format

```r
fisher.test(table(dat$variable1, dat$variable2))
```

