---
title: Clean Date
---

```{r setup, include=FALSE}
here::i_am("clean/date/clean-date.qmd")
library(here)
```

```{r}
library(lubridate)
```


## Format Date

Format date to `date` class.

```{r fmt_date_custom}
fmt_date_custom <- function(x) {
    out <- if (inherits(x, "POSIXt")) {
        lubridate::as_date(x)
    } else if (is.character(x)) {
        dmy(x)
    } else {
        NA
    }
    out
}
```

**Example:**

```{r}
fmt_date_custom("01/01/2020")
fmt_date_custom(Sys.Date())
```

## Time Interval

### Using `time_length()`

To convert a `difftime` object to **month** or **year intervals**, you can use the `lubridate` package, which provides functions like `time_length()` and `as.period()` for handling time intervals.

```{r}
library(lubridate)

# Create difftime object
start_date <- dmy("12/11/2001")
end_date <- dmy("12/12/2003")
time_diff <- end_date - start_date

# Convert to months
months_diff <- time_length(as.duration(time_diff), "months")

# Convert to years
years_diff <- time_length(as.duration(time_diff), "years")

# Print the results
print(paste("Months:", months_diff))
print(paste("Years:", years_diff))
```

**Explanation:**

1. **`time_length()`**:
   - Converts a `duration` object into the specified unit (e.g., months, years).
   
2. **`as.duration(time_diff)`**:
   - Converts the `difftime` object into a `duration` object, which is required by `time_length()`.

3. **Units**:
   - `"months"`: Converts the interval into months, assuming an average month length of ~30.44 days.
   - `"years"`: Converts the interval into years, assuming 365.25 days per year.


### Using `interval()` 


If you need to calculate **exact calendar intervals** (e.g., the number of full months or years between two dates), use `interval` and `int_length`:

```{r}
# Exact calendar months
months_diff_exact <- interval(start_date, end_date) %/% months(1)

# Exact calendar years
years_diff_exact <- interval(start_date, end_date) %/% years(1)

# Print the results
print(paste("Exact months:", months_diff_exact))
print(paste("Exact years:", years_diff_exact))
```
